---
title: "Final Paper"
author: "Kristina Becvar"
date: "8/13/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(blogbuilder)
library(distill)
library(dplyr)
library(cowplot)
library(Hmisc)
library(viridis)
knitr::opts_chunk$set(echo = TRUE)
```

## **Introduction**

*Deﬁne the research question or problem you seek to address. What animated your concern with this research question? What are you arguing?*

I believe that there is a tendency for companies to place and governments to approving zoning for pollution-generating facilities. This is not a novel argument, and I have been passionate about this concern for a long time. I am particularly concerned about the town where I grew up in Illinois, so I decided to look at environmental justice data for the state.

Communities of color, which are often also poor, have historically been specifically targeted to host facilities that have negative environmental impacts. I believe that despite our claims as a society to be champions of equality and justice, there is still that target on both majority minority communities and low-income communities as compared to the general population.

## **Data**

*Describe your data source. Why is this an appropriate choice? What are the variables that you are speciﬁcally interested in? Describe those variables, both numerically and visually.*

**The data I used is from the EPA mapping tool "EJSCREEN", an environmental justice screening tool that has mapping functions to see the locations of pollution and which communities are affected the most by that pollution.**

The data is gathered from multiple primary sources including the EPA federal and regional studies, CDC reporting, and state work. The EPA site allows users to download the dataset in a national summary format as well as by state, which is the dataset I am using in this analysis. In order to make the data more manageable, I am focusing only on my state's data - Illinois.

From reading the documentation provided with the dataset by the EPA, I know that the variables consist of both environmental data and demographic data.

**Environmental Variables**

Environmental variables tracked in this dataset primarily include calculations representing the proximity of census tracts and population blocks to certain environmental risks. These risks include sites that are on the National Priorities List ("Superfund" sites), facilities that have to file a Risk Management Plan ("RMP") because they produce toxic substances or emit toxic byproducts, facilities that are hazardous waste treatment, storage or disposal sites ("TSDF" sites), levels of diesel particulate matter in the air, traffic proximity and volume, particulate matter <2.5, ozone level in air, and index indicators for air toxics cancer risk, air toxics respiratory hazards, and proximity to major direct dischargers to water.

I have focused on looking at the first three environmental variables and named them as:

+ "superfund.area", the variable indicating the count of proposed and listed "Superfund" sites within 5 km of a population, divided by the distance away in kilometers.

+ "rmp.area", the variable indicating the count of facilities that are RMP sites within 5 km of a population, divided by the distance away in kilometers.

+ "tsdf.area", the variable indicating the count of facilities that are TSDF sites within 5 km of a population, divided by the distance away in kilometers.

**Demographic Variables**

Demographic variables tracked in this dataset include the total population of census tracts and population blocks. In addition, it gives calculations representing proportional data for each tract or block for the percentage of: people of color, low-income-defined as no more than twice the federal poverty level, education below high school, in linguistic isolation, under age 5, over age 64, and in pre-1960 housing. There is also a demographic index, which is based on 2 of those factors, percentage of low-income and people of color.

+ "census.region" is the geographical census tract being measured, which can include multiple population blocks being studied by the EJSCREEN data.

+ "population" is the count of persons living in the correlating census tract.

+ "dem.index" represents a particular census tract's proportion of minority population plus the percentage of low-income population divided by 2. 

There is a large amount of map color data and map text included in this dataset. There are multiple location variables primarily divided by a unique block with defined features by coordinate. These blocks are also categories by by EPA region, state, and census FIPS code area. I am not analyzing this additional data in this project.

```{r echo=TRUE}
ejscreen <- read.delim("../data/EJSCREEN_2020_IL.csv", sep = ",", header  = TRUE)
```

```{r echo=TRUE}
ejscreen.tidy <- as_tibble(ejscreen)
```

**Describing Variables**

I have taken the census region and changed the format to scientific notation before taking the given variable names and renaming them to easily identifiable names as listed below. Finally, I group the databy census tract, and calculate the remaining variables by the mean of the population blocks I am combining in the process.

+ PNPL = superfund
+ PRMP = rmp.area
+ PTSDF = tsdf.area
+ ID = census.region
+ ACSTOTPOP = population
+ VULEOPCT = dem.index

```{r echo=TRUE}
options(scipen = 999)
(ejscreen.clean <- ejscreen.tidy %>% 
   rename(superfund.area = PNPL,
          rmp.area = PRMP,
          tsdf.area = PTSDF,
          census.region = ID,
          population = ACSTOTPOP, 
          dem.index = VULEOPCT) %>% 
   select(superfund.area, rmp.area, tsdf.area,  census.region, population, dem.index))
```

```{r echo=TRUE}
ejscreen.mean <- ejscreen.clean %>%
  group_by(census.region) %>%
  summarise(mean.rmp = mean(rmp.area), 
            mean.superfund = mean(superfund.area),
            mean.tsdf = mean(tsdf.area),
            mean.demindex = mean(dem.index),
            mean.population = mean(population))
ejscreen.mean
```

**Analyzing the Relationships Between Variables**

I can create a correlation matrix that gives an overview of the correlations for all combinations of two variables, rounded to 2 decimals. This argument produces Pearson correlations by default.

```{r echo=TRUE}
round(cor(ejscreen.mean),
  digits = 2 
)
```

After analyzing the Pearson correlations, I can see that there are some relationships with stronger positive linear relationships than others. Those moderately strong positive correlations (0.51 to 0.7) are: 

+ Mean Demographic Index to Mean RMP Area Frequency
+ Mean Demographic Index to Mean TSDF Area Frequency

The weak positive correlations (0.3 to 0.5) are:

+ Mean Demographic Index to Mean Superfund Area Frequency
+ Mean TSDF Area Frequency to Mean RMP Area Frequency

**Making a Null and Alternative Hypothesis**

H0: ρ=0 (meaning that there is no linear relationship between the two variables)

H1: ρ≠0 (meaning that there is a linear relationship between the two variables)

I used cor.test to determine the p-value to examine the relationships between the EJSCREEN mean demographic index score for a given census tract and the mean number of environmental threats in those areas as measured by the mean count of these sites in the mean.rmp, mean.superfund and mean.tsdf variables.

I then need to look at the p-values of each relationship in my data set using the 'Hmisc package'. This shows each p-value rounded to three decimals.
 
```{r echo=TRUE}
res <- rcorr(as.matrix(ejscreen.mean))
round(res$P, 3)
```

After analyzing the matrix, it is clear that we do have some correlations with statistically significant correlations using a significance level of α=0.05

P-value ≤ α: The correlation is statistically significant

+ P-values for mean.rmp and mean.tsdf
+ P-values for mean.population and mean.tsdf
+ P-values for mean.demindex and mean.rmp, mean.superfund, and mean.tsdf

P-value > α: All other non-categorical relationships

**Summary of Initial Variable Analysis**

At first review of my initial statistical analysis of my hypothesis, it does seem that there are statistically significant, positive correlations between increasing demographic index ratings and the frequency of all three environmental risk variables I have examined. 

Looking further, the strongest relationships are between the demographic index ratings and the frequency of TSDF sites and RMP sites within 5 km of the census region. There is a significant but weaker relationship between the demographic index ratings and the frequency of the Superfund sites within 5 km of the census region. 

This could be explained logically by the nature of the environmental risk; Superfund sites are those that have been identified as former polluting sites that have urgent for remediation, which is more likely to be in areas where there are fewer vulnerable and low-income populations and more vocal and high-influence populations demanding urgency to the remediation.

On the other hand, the TSDF sites (toxic storage and disposal facilities) and RMP sites (facilities that emit toxic pollutants that require RMP (risk management plans) on file with the EPA) are those that are actively operating in vulnerable populations.

```{r echo=TRUE}
Demindex.Percentile_00 <- min(ejscreen.mean$mean.demindex)
Demindex.Percentile_25 <- quantile(ejscreen.mean$mean.demindex, 0.25)
Demindex.Percentile_50 <- quantile(ejscreen.mean$mean.demindex, 0.50)
Demindex.Percentile_75 <- quantile(ejscreen.mean$mean.demindex, 0.75)
Demindex.Percentile_100 <- max(ejscreen.mean$mean.demindex)

RB = rbind(Demindex.Percentile_00, Demindex.Percentile_25, Demindex.Percentile_50, Demindex.Percentile_75, Demindex.Percentile_100)

dimnames(RB)[[2]] = "Value"

RB
```

```{r echo=TRUE}
ejscreen.mean$Demindex.Quartile[ejscreen.mean$mean.demindex >= Demindex.Percentile_00 & ejscreen.mean$mean.demindex <  Demindex.Percentile_25]  = "1st Quartile"
ejscreen.mean$Demindex.Quartile[ejscreen.mean$mean.demindex > Demindex.Percentile_25 & ejscreen.mean$mean.demindex <  Demindex.Percentile_50]  = "2nd Quartile"
ejscreen.mean$Demindex.Quartile[ejscreen.mean$mean.demindex > Demindex.Percentile_50 & ejscreen.mean$mean.demindex < Demindex.Percentile_75] = "3rd Quartile"
ejscreen.mean$Demindex.Quartile[ejscreen.mean$mean.demindex > Demindex.Percentile_75 & ejscreen.mean$mean.demindex <=  Demindex.Percentile_100]  = "4th Quartile"

ejscreen.mean
```

## **Visualization**

*Develop a visualization that best addresses your research question. Describe the visualization in detail. Then, explain what other options you considered for visualizing the data, and explain why the option you chose was the best option.*

**Scatterplot 1**

I created a faceted scatterplot showing the mean population of each census area on the x axis and the mean frequency of each census area for each environmental risk on the y axis. Each quartile of the demographic index rating is illustrated independently, showing a visual acknowledgement of what the statistical analysis showed.

```{r echo=TRUE}
(ggplot(data = ejscreen.mean) + 
  geom_point(mapping = aes(x = mean.population, y = mean.tsdf)) + 
  facet_wrap(~ Demindex.Quartile, nrow = 2))
(ggplot(data = ejscreen.mean) + 
  geom_point(mapping = aes(x = mean.population, y = mean.superfund)) + 
  facet_wrap(~ Demindex.Quartile, nrow = 2))
(ggplot(data = ejscreen.mean) + 
  geom_point(mapping = aes(x = mean.population, y = mean.rmp)) + 
  facet_wrap(~ Demindex.Quartile, nrow = 2))
```
**Scatterplot 2**

The most significant correlation looks like the relationship between the mean number of TSDF sites and a higher demographic index. I want to visualize this differently, so I'm using color to look at the demographic index quartiles against the scatterplot of population and mean TSDF sites. It displays a visual representation of how many more TSDF sites are truly located in the highest minority and low-income population areas.

```{r echo=TRUE}
v <- ggplot(ejscreen.mean, aes(mean.population, mean.tsdf))
v + geom_point(size=4, aes(colour = Demindex.Quartile)) +
    scale_color_viridis(discrete=TRUE) +
    theme_bw()
```

**Scatterplot 3**

The last scatterplot I tried displays a representation of the mean population of a census tract and the mean number of TSDF areas with a line smoothing the points to highlight the trajectory of the data. 

```{r echo=TRUE}
ggplot(ejscreen.mean, aes(x = mean.demindex, y = mean.tsdf)) + 
  geom_point() +
  geom_smooth(color = "purple") +
  theme_half_open(12)
```

**Density Plot**

I also used the package "cowplot" and visualized the Demographic Index Quartile densities against each of the environmental risk factor frequencies. This visualization shows relationships between the patterns graphically well, but primarily, again, for the strongest correlating variable;the mean TSDF rating and areas in the 4th quartile of the demographic index.

```{r echo=TRUE}
ggplot(ejscreen.mean, aes(mean.tsdf, fill = Demindex.Quartile)) + 
  geom_density(alpha = 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_minimal_hgrid(12)
```

```{r echo=TRUE}
ggplot(ejscreen.mean, aes(mean.rmp, fill = Demindex.Quartile)) + 
  geom_density(alpha = 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_minimal_hgrid(12)
```

```{r echo=TRUE}
ggplot(ejscreen.mean, aes(mean.superfund, fill = Demindex.Quartile)) + 
  geom_density(alpha = 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_minimal_hgrid(12)
```

**Visualization Choices**

I considered using bar graphs or pie charts, but having two continuous variables limited what I could demonstrate using them. In order to use a bar graph, I would need to do something like making them categorial values to reflect quartiles. I don't believe that would be as accurate of a representation as a scatterplot, which captures more points of data. 

## **Reﬂection** 

*Describe your experience with the process. What decisions in your pipeline are you most concerned about potentially inﬂuencing your ﬁndings? What were the most challenging and time-consuming aspects of the project? What do you wish you had been able to do? If you were to continue the project, what would your next steps be?*

I found this project engaging and the process somewhat addictive. I started down more paths than I have documented in this final paper but have run out of work through them completely. I plan to continue down those paths as diving into these concepts has been very interesting, and I am finding myself more and more competent in the language of R.

I am most concerned about my choice in taking the demographic index score and dividing it into quartiles. If I were to continue the project, I would definitely work with the variables to interpret more of them into different categorical units. I would also go back and look at more variables representing different environmental and demographic data. Specifically, I wish I had been able to analyze the traffic data. In my hometown, intermodal centers have taken over the landscape to the point that the county is now officially the largest inland port in North America. The intermodal centers take advantage of the county being where two interstates meet, which is also adjacent to three different rivers and a major train hub. Unfortunately, most of the residents did not want any of these intermodal centers to build in the county, but being a low income area, the residents lacked the power or resources to challenge the developments. I believe the sharp rise in asthma among area children could correlate with the increse in diesel truck traffic. There are limitations to this dataset in making that analysis though, so I would need to bring in additional data to complete that study.

## **Conclusion** 

*Discuss your research question again brieﬂy and state succinctly the preliminary ﬁndings. Also discuss the potential ramiﬁcations of action – or inaction – on this policy.*

There is definitely a statistically significant correlation between census tracts with the highest proportion of minority and low income populations as calculated in the demographic index variable and the environmental risks I analyzed. These preliminary findings lend credence to the central question of my project, which was whether there is a tendency for companies to place and governments to approve zoning for pollution-generating facilities in those vulnerable communities.

## **Bibliography:** 

*United States Environmental Protection Agency (EPA), 2020 version, updated July 1, 2021. EJSCREEN. Retrieved: July 16, 2021 (https://gaftp.epa.gov/EJSCREEN/2020/)*
